---
title: 'Statistical report: Association between ward type and outcomes in patients with heart failure'
author: 'Statistician: Lina Benson'
  
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
    number_sections: yes
link-citations: yes
bibliography: references.bib
nocite: '@*'
---

\newpage 
\tableofcontents 
\listoffigures
\listoftables
\newpage


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, include = TRUE, comment = "",
  warning = FALSE, message = FALSE, fig.pos = "H",
  fig.path = "../output/figs/"
)
options(knitr.kable.NA = "")
```

```{r adjust_directory, include=FALSE}
# since report is placed in subdirectory of project
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

```{r load_project}
# 1. Set options in config/global.dcf
# 2. Load packages listed in config/global.dcf
# 3. Import functions and code in lib directory
# 4. Load cache 

ProjectTemplate::reload.project()

# variables that should be presented in tables and included in models
source("./src/vars_for_tabs_mods.R")
```             

# Data handling

## Data source

Dataset prepared by Bodil Svennblad with corrections version 6 (rs.data6) with 
end follow-up 2016-12-31 consisting of data from SwedeHF, SCB (LISA Registry) 
and Socialstyrelsen (National Patient Registry and Cause of Death Registry). 
Note that ward type is no longer collected in SwedeHF due to problems with the 
definition and therefore data from the new SwedeHF is not included in the analysis. 
This also means that some care should be taken when interpreting the results. 

## Created variables 

Variables prepared in rs.data6 are documented elsewhere. This section specifies 
additional recoding and derivation of variables for the purpose of current 
analysis that are not considered to be self-explanatory. 

Income is considered below median if it is below the 50 percentile for that 
respective year of inclusion, and above median if it is above or equal to the 50 percentile. 

NT-proBNP is considered below median if it is below the 50 percentile 
and above median if it is above or equal to the 50 percentile. 

ras is defined as yes if arb, acei or arni = yes (and arb and acei both 
non-missing) and no if both arb and acei = no. Since arni was introduced 
into SwedeHF at a later date all missing are presumed to be no.   

Atrial_fib_flutter, IHD, Diabetes, Hypertension and valve disease are 
combined from registrations in SwedeHF and the derived variables from 
the National patient registry as follows: 

d_Atrial_fib_flutter is yes if any of the following are fulfilled: 
com_AF = "yes" or fibrillation = "yes" or ekg = "Atrial fibrillation",

d_IHD is yes if any of the following are fulfilled: 
com_IHD = "yes" or revascularized = "yes" or com_pci = "yes" or com_cabg = "yes 
(there is approx 70% missing in ihd in SwedeHF registations post 2013 so 
this variable was not included in the combinded variable),

d_Diabetes is yes if any of the following are fulfilled: 
com_DM = "yes" or diabetes = "yes", 

d_Hypertension is yes if any of the following are fulfilled: 
com_hypertension = "yes" or hypertension = "yes", 

d_Valvedisease is yes if any of the following are fulfilled: com_Valvular = "yes" 
or valveDisease = "yes" or heartValveSurgery = "yes"

If the variable in SwedeHF is missing (no missing in the National patient 
registry) the combined variable is also set to missing. 

Observe that index date is for this project defined as admission NOT discharge 
for in-patient visits. So time to death/hospitalization is from admission/visit 
to event/censoring or death. 

## Inclusion/exclusion criteria

```{r flow}
mykable(flow, caption = "Flowchart")
```

N unique patients in (). Observe that one patient can have multiple posts and 
therefore be counted more than once on each row (so for example can be in both the
excl cardiology cell and the excl medicine cell with different posts). 

First patient in: `r min(wdata$date)` and last patient in: `r max(wdata$date)`.  

# Statistical analysis 

## General

All analyses were performed using `r sessionInfo()$R.version$version.string` [@r]. 
The level of significance is set to 5%, two-sided. No adjustment for multiple 
comparisons were made and therefore the results should be viewed with some care.

## Missing data

Missing data was imputed with multiple imputation using mice [@mice] for 10 
datasets and 10 iterations. Variables included in the model are indicated in 
Table \ref{tab:tab1overall}. The primary outcome was included as the Nelson-Aalen estimator. 
Ward type is not included in the imputation model.  

## Time plot

```{r, child = "../src/barplot_cardio.Rmd"}

```

\clearpage
\newpage

## Baseline characteristics and Odds ratios
  
Categorical variables are presented with n (%) and tested with chi-square test 
and continuous variables with median (q1-q3) and tested with Mann-Whitney U test. 
Odds ratios are estimated using logistic regression. Note that the p-values 
for the descriptive statistics should be redundant when presenting p-values 
from the logistic regression in the same table. 

Variables were chosen based on clinical relevance. BMI was not included in 
model due to the very percentage of missing observations.  

```{r, child = "../src/tab1.Rmd"}

```

### Assumptions

Outliers were investigated with Cook's distance and multicollinearity 
with the variance inflation factor in the full model in the overall population. 
No action was deemed necessary.    

## Outcomes

Summary of follow-up time (years, days):
```{r}
summary(wdata$timeTodeath) / 365
summary(wdata$timeTodeath)
```

### KM curves

Crude time to the respective outcomes was presented using the Kaplan–Meier analysis
and differences between ward type tested with the log-rank test. 

```{r, child = "../src/km.Rmd"}

```

\clearpage
\newpage

### Models

Incidence per 1000 py was calculated with 95% Poisson confidence intervals. 

A cox proportional hazards regression was used to evaluate the association 
between ward and the respective outcomes. Data were censored at 1 year follow-up 
or death, so this assumes that the risk of cv mortality and hf hospitalization 
would have been the same for a censored patient had the censoring not occurred. 
Variables adjusted for in the model were chosen according to clinical relevance. 
Currently the same variables are adjusted for as are included in the logistic 
regression. 

In-hospital mortality for the in-patient population was analyzed with logistic 
regression adjusting for the same variables as above. 

```{r, child = "../src/outtab.Rmd"}

```

\newpage

#### Subgroups

Possible differences over subgroups were investigated by including the 
interaction between the respective variable and clinic. 
Note that NYHA is dichotomized in the subgroup analysis whereas 
included as the original variable in the main model. 

```{r, child = "../src/forest.Rmd"}

```

\newpage

#### Assumptions

The assumptions of the models were checked for the primary endpoint all-cause mortality
in the adjusted model and the overall population. The proportional hazards assumption 
was investigated using the scaled Schoenfeld residuals (cox.zph in [@survival-package]). 
Possible outliers were visually inspected by plotting the dfbetas. 
Age, location and systolic bp and ras were deemed not to have proportional hazards 
and were therefore included as strata variables in a sensitivity analysis for 
the primary endpoint.  

# Reproducibility

## R session information {#sessioninfo}

`r sessionInfo()$R.version$version.string` [@r] is used, for more details see below.

```{r sessinfo}
sessionInfo()
```

## R code

The R code for all data handling and statistical analyses are found: 
https://github.com/linabe/Wardtype. On publication the repository is made 
public under https://github.com/orgs/KIHeartFailure/wardtype_outcomes 
and it is therefore possible to link to it from the resulting article 
for increased transparency and code sharing. No data or output is stored in the 
repository. 

# References
